Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-12 paper.qmd
Start updating with new quollr version

```{r}
#| label: two-curvy-training
training_data_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_data.rds")
```

@@

data_two_curvy <- training_data_two_curvy |>
  mutate(type = "data")

@@

<!--tSNE applied for Two curvy data-->
```{r}
#| label: tsne-two-curvy
tsne_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds") 
```

@@

algo_obj_two_curvy <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  num_x_bins = num_bins_x_two_curvy)

tsne_two_curvy_scaled <- algo_obj_two_curvy$nldr_scaled
tr_from_to_df_two_curvy <- algo_obj_two_curvy$tr_from_to_df
df_bin_centroids_two_curvy <- algo_obj_two_curvy$df_bin_centroids
df_bin_two_curvy <- algo_obj_two_curvy$df_bin
hex_grid_with_counts <- algo_obj_two_curvy$hex_grid_with_counts
hex_grid_nonempty <- algo_obj_two_curvy$hex_grid_nonempty
bin_width <- algo_obj_two_curvy$a1

@@

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = tSNE1, y = tSNE2)) +
  geom_point(alpha = 0.5, color = clr_choice, size = 0.5)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = tSNE1, y = tSNE2), 
             alpha = 0.5, size = 0.5, color = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

df_bin_centroids_two_curvy <- df_bin_centroids_two_curvy |>
  dplyr::filter(std_counts > 0)

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

training_data_two_curvy <- training_data_two_curvy |>
  mutate(ID = row_number())

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = sqrt_row_wise_total_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

@@

df_bin_two_curvy <- df_bin_two_curvy |>
  select(-hb_id) |>
  mutate(type = "model")

@@

<!-- df_bin_two_curvy <- df_bin_two_curvy |> -->
<!--   select(-hb_id) |> -->
<!--   mutate(type = "model") -->

@@

df_b_two_curvy4 <- df_bin_two_curvy4 |>
  dplyr::filter(hb_id %in% df_bin_centroids4$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_two_curvy4 <- df_b_two_curvy4[match(df_bin_centroids4$hexID, df_b_two_curvy4$hb_id),] |>
  dplyr::select(-hb_id)

@@

df_b <- df_bin_gau1 |>
  dplyr::filter(hb_id %in% df_bin_centroids_gau1$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)

@@

df_bin_dens_clust <- df_bin_dens_clust |>
  select(-hb_id) |>
  mutate(type = "model")

@@

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hb_id %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hb_id),] |>
  dplyr::select(-hb_id)

@@

df_b_mnist <- df_bin_mnist |>
  dplyr::filter(hb_id %in% hexID_mnist) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_mnist <- df_b_mnist[match(hexID_mnist, df_b_mnist$hb_id),] |>
  dplyr::select(-hb_id)

>>>>

```{r}
#| label: two-curvy-training
training_data_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_data.rds") |>
  mutate(ID = row_number())
```

@@

data_two_curvy <- training_data_two_curvy |>
  select(-ID) |>
  mutate(type = "data")

@@

<!--tSNE applied for Two curvy data-->
```{r}
#| label: tsne-two-curvy
tsne_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds") |>
  mutate(ID = row_number()) |>
  rename(c("emb1" = "tSNE1",
           "emb2" = "tSNE2"))
```

@@

algo_obj_two_curvy <- fit_highd_model(
  highd_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  bin1 = num_bins_x_two_curvy, 
  q = 0.1, 
  benchmark_highdens = 5)

tsne_two_curvy_scaled <- algo_obj_two_curvy$nldr_obj$scaled_nldr
tr_from_to_df_two_curvy <- algo_obj_two_curvy$trimesh_data
df_bin_centroids_two_curvy <- algo_obj_two_curvy$model_2d
df_bin_two_curvy <- algo_obj_two_curvy$model_highd
hex_grid <- algo_obj_two_curvy$hb_obj$hex_poly
counts_df <- algo_obj_two_curvy$hb_obj$std_cts

hex_grid_with_counts <- left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hexID"))

hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids_two_curvy$hexID)

bin_width <- algo_obj_two_curvy$hb_obj$a1

@@

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, color = clr_choice, size = 0.5)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, color = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

# df_bin_centroids_two_curvy <- df_bin_centroids_two_curvy |>
#   dplyr::filter(std_counts > 0)

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = sqrt_row_wise_total_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

@@

df_bin_two_curvy <- df_bin_two_curvy |>
  select(-hexID) |>
  mutate(type = "model")

@@

<!-- df_bin_two_curvy <- df_bin_two_curvy |> -->
<!--   select(-hexID) |> -->
<!--   mutate(type = "model") -->

@@

df_b_two_curvy4 <- df_bin_two_curvy4 |>
  dplyr::filter(hexID %in% df_bin_centroids4$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_two_curvy4 <- df_b_two_curvy4[match(df_bin_centroids4$hexID, df_b_two_curvy4$hexID),] |>
  dplyr::select(-hexID)

@@

df_b <- df_bin_gau1 |>
  dplyr::filter(hexID %in% df_bin_centroids_gau1$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hexID),] |>
  dplyr::select(-hexID)

@@

df_bin_dens_clust <- df_bin_dens_clust |>
  select(-hexID) |>
  mutate(type = "model")

@@

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hexID %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hexID),] |>
  dplyr::select(-hexID)

@@

df_b_mnist <- df_bin_mnist |>
  dplyr::filter(hexID %in% hexID_mnist) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_mnist <- df_b_mnist[match(hexID_mnist, df_b_mnist$hexID),] |>
  dplyr::select(-hexID)
